---
title: "Practical Machine Learning - Course Project"
author: "Samer Araabi"
date: "July 15, 2015"
output: html_document
---

## Overview

This following analysis is designed to use machine learning models to predict the manner in which participants of the Human Activity Recognition project performed a series of predetermined  exercises. After cleaning the dataset to isolate more robust variables, a "random forest" machine-learning model was able to replciate the system used by Groupware@LES with over 99% accuracy.

More information on the original study can be accessed at the following URL: http://groupware.les.inf.puc-rio.br/har

## Retrieving and storing the data

The data was pulled directly from the online HAR archive and loaded into R using the following code.
```{r}
train_url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
test_url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
download.file(train_url, dest="training.csv", method="curl")
download.file(test_url, dest="testing.csv", method="curl")
training <- read.csv("training.csv")
testing <- read.csv("testing.csv")
```

## Cleaning and formatting

A preliminary analysis of the data revealed that many of the columns had little to no data.

```{r}
nas <- list()
for(i in names(training)){
  nas <- c(nas, sum(is.na(training[i])))
  }
sum(nas>10000)
```

These were removed from the dataset, along with columns that were blank:

```{r}
clean_train <- training[nas==0]
clean_test <- testing[nas==0]

empty <- list()
for(i in names(clean_train)){
  empty <- c(empty, sum(clean_train[i]==""))
  }

clean_train <- clean_train[empty==0]
clean_test <- clean_test[empty==0]
```

The first seven columns, which contained descriptors but no predictive data, were also removed:

```{r}
clean_train <- clean_train[,-c(1:7)]
clean_test <- clean_test[,-c(1:7)]
```

Lastly, the remaining 52 variables (except for the dependent "classe" variable, were converted to numerics to facilitate the modeling:
```{r}
for(i in 1:52){
  clean_train[,i] <- as.numeric(clean_train[,i])
  }
str(clean_train)
```

## Partitioning
The cleaned data was then partitioned into training and testing groupings, with 70% in the training group and the remaining 30% in the testing group. Note that the "caret" package is required to perform the data partition, and that a seed was set to ensure reproducibility. 

```{r}
library(caret)
set.seed(6492)
inTrain = createDataPartition(y=clean_train$classe, p= 0.7, list=FALSE)
training_data <- clean_train[ inTrain,]
testing_data <- clean_train[-inTrain,]

```

## Modelling

The data was modeled using a "random forest" machine-learning method. Note that parallel processing was used to reduce runtime for the model-building:
```{r}
library(randomForest)
library(doParallel)
registerDoParallel(cores=2)
modFit <- train(classe ~ ., method="rf", data=training_data)
modFit$finalModel
```

## Cross-Validation

The model was validated on the testing data to determine the out-of-sample error rate, which was less than 1%:

```{r}
mean(predict(modFit, testing_data) == testing_data$classe)
```

We would therefore expect to accurately predict the manner of a participant's exercises with approximately 99% accuracy. 